\documentclass{article}
\usepackage[utf8]{inputenc}

\title{Purchase: How Explainable are Simplified Graph Neural Networks}
\author{Skye Purchase}

\begin{document}
\maketitle

Name: Skye Purchase 

College: Peterhouse

CRSID: atp45

Director of Studies: Robert D. Mullins

\section{Current Project Ideas}

My project will focus on implementing two papers; Simplified Graph Convolutions (SGC) and the second Graph Convolution Explainer (GCExplainer). To see if the results from GCExplainer on SGC are more meaningful than those from standard Graph Neural Network (GNN). The project thus has two phases:

\begin{itemize}
    \item Implement SGC and run on GCExplainer datasets.
    \item Implement GCExplainer and test on SGC and compare results.
\end{itemize}

My current success metric is to implement GCExplainer on SGC and evaluate it against GCExplainer on standard GNNs.

I will need to research the techniques used in these papers and how they are implemented primarily reading the paper and related work.

\section{Supervisors}

\begin{itemize}
    \item Lucie Charlotte Magister - agreed and met
    \item Pietro Babiero - agreed and met
    \item (Pietro Lios - agreed and met)
\end{itemize}

I have contacted Pietro Lios who put me in touch with my supervisors. I have had a preliminary meeting with both my supervisors and we have scheduled regular meetings for the future.

\section{Computing Resources}

High performance cluster (HPC) as I will require a large number of GPU hours and speed when training and testing my models. 

Local CL server access (idun, heimdall, etc) for less costly iterations and testing of GPU models rather than using costly HPC time.

My own machine for remote work on the project but will hopefully be focusing on ssh into these other services.

Based on which datasets I use I will likely need extra storage to save these. Further to this storage space on the servers would be used when loading up the models to train.

Would require other software packages such as PyTorch and PyTorch Geometric to help with the framework of building, testing and training my models

\end{document}